RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-11-06 introduction.tex
MASTERFIX

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, goal models for many complex, real-world information systems -- e.g., air-traffic management systems, systems that support industrial production processes, or government and healthcare services -- are not constructed in a short amount of time, but rather over the course of several workshops with stakeholders and requirements engineers.  %SG: I reword and add the following sentences as the paragraphs below are very disjunct.
Thus, developing goal models for a complex and large system is not a trivial task and can be very cumbersome. In such situations, failure in recording the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development.

The first challenge with the goal modeling phase, particularly in large projects, is related to their dynamic nature: goal models continuously change and evolve. Stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Stakeholders may change their opinion about a modeling decision in between two modeling sessions, which may require revisions of a goal model. If the rationales behind these revisions are not properly documented, alternative ideas and opposing views that could potentially lead to different goal models might be lost, as the resulting goal model only shows the end product of a long process and not the discussions during the modeling process. Furthermore, other stakeholders, such as developers who were not the original authors of the goal model, may have to make sense of a goal model in order to, for example, use it as an input in a later RE stage or in the development phase. If preferences, opinions and rationales behind the goal models are not stored explicitly, it may not only be more difficult to understand the model, but the other stakeholders may end up having similar discussions throughout the design and development phase as well.

@@

To summarize, what is needed is a way to record the rationales (beliefs, opinions, discussions, ideas) underlying a goal model. It should be possible to see how these rationales change during the goal modeling process, and they  should be clearly linked to the intentional elements of the resulting goal model. In order to do this, we propose a framework with its tool-support which combines traditional goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{BenchCaponDunne2007}. We have identified \textbf{five important requirements} for our framework:

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying argumentation on the goal model, and vice versa.
\item 
There must be a methodology for the framework to guide the practitioners in its application in real cases.
\item 
The framework must have a tool-support.
\end{enumerate}

@@

Following from our previous work~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we develop our \emph{RationalGRL} framework, which combines the Goal-oriented Requirements Language (GRL) \cite{Amyot:2010:EGM:1841349.1841356} with a technique from argumentation theory called \emph{argument schemes} (or argumentation schemes~\cite{walton-etal2008}). The \emph{RationalGRL} framework includes argument schemes and their critical questions, a meta-model, a tool-support and a methodology on how to develop such framework.

Argument schemes are reusable patterns of reasoning that capture the common approaches in which humans argue and reason. \emph{Critical questions} which are associated with argument schemes, point to typical sources of doubt or implicit assumptions people make when raising an arguement. Argument schemes are a good choice for modeling discussions about a goal model, as they can guide users in systematically deriving conclusions and making assumptions explicit~\cite{murukannaiah2015}.

One argument scheme that is important when reasoning about goals is the argument scheme for practical reasoning~\cite{walton1990,atkinson2007}, which has been used for, among other things, modeling software design discussions~\cite{BlackEtal2013}. Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyze and guide stakeholders' discussions about goal models. Our approach, thus, provides a rationalization to the elements of the goal model in terms of underlying arguments, and helps in understanding why parts of the model have been accepted while others have been rejected. Our list of argument schemes was constructed by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. This ensures that the argument schemes we propose are close to actual real-world discussions the stakeholders have (\textbf{requirement 1}).

%SG: Need to come back to this later.
The meta-model of the RationalGRL framework clearly specifies the traceability links between the arguments based on the schemes and the GRL models (\textbf{requirement 2}). In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework~\cite{Dung1995}. We, then, formally capture the link between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models. These formal traceability links allow us to compute the effect of the arguments and counterarguments proposed in a discussion on a GRL model (\textbf{requirement 3}). In other words, we can determine whether the elements of a GRL model are acceptable given potentially contradictory opinions of stakeholders. Thus, we add a new formal evaluation technique for goal models that allows us to assess the \emph{acceptability} of elements of a goal model (in addition to their \emph{satisfiability}~\cite{Amyot:2010:EGM:1841349.1841356}).

One of our major goals is to provide means for RE practitioners to capture the underlying arguments of goal models by using our RationalGRL framework (\textbf{requirement 4}). Thus, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. To show the applicability of the RationalGRL methodology, we illustrate its steps with our traffic simulator case study. Finally, we have implemented the Rational-GRL tool-support, a web-based prototype\footnote{\url{http://www.rationalgrl.com}}, for modeling argument schemes and critical questions and for reasoning about goal models with respect to them (\textbf{requirement 5}).

%SG: Need to reword this later.
The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study, an in-depth explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL language metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question that is applied. Section~\ref{sect:methodology} discusses the RationalGRL methodology and a discussion of the prototype tool we have developed for RationalGRL. Finally, Section~\ref{sect:discussion} covers related work, future work, and a conclusion.

>>>>

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, goal models for many complex, real-world information systems -- e.g., air-traffic management systems, systems that support industrial production processes, or government and healthcare services -- are not constructed in a short amount of time, but rather over the course of several workshops with stakeholders and requirements engineers.  %SG: I reword and add the following sentences as the paragraphs below are very disjunct.
Developing goal models for such complex and large systems is not a trivial task and can be very cumbersome. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development.

The first challenge with the goal modeling phase, particularly in large projects, is related to their dynamic nature: goal models continuously change and evolve. Stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Stakeholders may change their opinions about a modeling decision in between two modeling sessions, which may require revisions of a goal model. If the rationales behind these revisions are not properly documented, alternative ideas and opposing views that could potentially lead to different goal models might be lost, as the resulting goal model only shows the end product of a long process and not the discussions during the modeling process. Furthermore, other stakeholders, such as developers who were not the original authors of the goal model, may have to make sense of a goal model in order to, for example, use it as an input in a later RE stage or in the development phase. If preferences, opinions and rationales behind the goal models are not stored explicitly, it may not only be more difficult to understand the model, but the other stakeholders may end up having similar discussions throughout the design and development phase as well.

@@

To summarize, what is needed is a way to record the rationales (beliefs, opinions, discussions, ideas) underlying a goal model. It should be possible to see how these rationales change during the goal modeling process, and they  should be clearly linked to the intentional elements of the resulting goal model. In order to do this, we propose a framework with tool support which combines traditional goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{BenchCaponDunne2007}. We have identified \textbf{five important requirements} for our framework:

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying argumentation on the goal model, and vice versa.
\item 
A methodology must be identified that can guide practitioners in using the framework.
\item 
The framework must have a tool support.
\end{enumerate}

@@

%MvZ: I shortened the paragraph below and put the full version in the conclusion. I think it makes more sense to reflect on the requirements in the conclusion, and to only give a brief overview of our framework in the introduction. Too much detail about argument schemes etc may confuse the reader.

Following from our previous work~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we develop the \emph{RationalGRL} framework, which extends an existing framework for goal modeling, the Goal-oriented Requirements Language (GRL) \cite{Amyot:2010:EGM:1841349.1841356}, with a technique from argumentation theory called \emph{argument schemes} (or argumentation schemes~\cite{walton-etal2008}). Argument schemes are reusable patterns of reasoning that capture the common approaches in which humans argue and reason. We develop a list of argument schemes by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. In order to specify clearly in what way RationalGRL extends GRL, we develop a metamodel, which specifies the traceability links between the arguments based on the schemes and the GRL models. In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework~\cite{Dung1995}. We, then, formally capture the link between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models.

In order to support practitioners in using the RationalGRL framework, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. We  implemented a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study and an explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and visa versa, and we develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question that is applied. Section~\ref{sect:methodology+tool} discusses the RationalGRL methodology and explain various features of the tool we developed. Finally, Section~\ref{sect:discussion} covers related work, future work, and a conclusion.

### 2017-11-06 main.tex
MASTERFIX

@@

\title{RationalGRL: A Framework for Argumentation and Goal Modeling}
\author{Marc van Zee \and Floris J. Bex \and Sepideh Ghanavati}
\institute{Marc van Zee \at Google Research, Z\"urich, Switzerland \\\email{marcvanzee@gmail.com} \and Floris Bex \at Department of Information and Computing Sciences, Utrecht University, Utrecht, The Netherlands \\\email{f.j.bex@uu.nl} \and Sepideh Ghanavati \at Department of Computer Science, Texas Tech University, Lubbock TX, USA \\\email{sepideh.ghanavati@ttu.edu}}

@@

\input{introduction}
\input{background}
\input{gmas}
\input{overview}
%\input{examples}
\input{formalframework}
\input{methodology}
%\input{tool}
\input{discussion}

>>>>

\newcommand{\rationalgrlurl}{\url{http://www.rationalgrl.com}}

@@

\title{RationalGRL: A Framework for Argumentation and Goal Modeling}
\author{Marc van Zee  \and Floris J. Bex\and Sepideh Ghanavati}
\institute{Marc van Zee \at Google Research, Z\"urich, Switzerland \\\email{marcvanzee@gmail.com} 
\and Floris Bex \at Department of Information and Computing Sciences, Utrecht University, Utrecht, The Netherlands \\\email{f.j.bex@uu.nl}
\and Sepideh Ghanavati \at Department of Computer Science, Texas Tech University, Lubbock TX, USA \\\email{sepideh.ghanavati@ttu.edu} }

@@

\input{introduction}
\input{background}
\input{casestudy}
\input{overview}
%\input{examples}
\input{formalframework}
\input{methodology}
%\input{tool}
\input{discussion}

### 2017-11-06 methodology.tex
MASTERFIX

\section{The RationalGRL Methodology and Tool}
\label{sect:methodology}

In previous sections, we have shown how the RationalGRL framework can capture stakeholder discussions, and how interactions between two types of reasoning, practical reasoning and goal modeling, leads to two interlinked models, RationalGRL and GRL models. In this section we clarify how practitioners can actually use the RationalGRL framework by proposing a methodology (\textbf{requirement 4}) and discussing a prototype RationalGRL tool (\textbf{requirement 5}).

@@

We propose the methodology shown in Figure~\ref{fig:rationalgrl-methodology} to develop a (Rational)GRL model, an earlier version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. %Here we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

@@

An important requirement of our framework is that it has tool support (\textbf{requirement 5}. This is for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, there does not yet exist any tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work, iv) having a light-weight web-based version of GRL is useful for the community in general. In this section we briefly highlight some features of the tool, and we explain some of its current limitations.

GRL has a well-documented and well-maintained tool called jUCMNav~\cite{jUCMNav}\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}, which is implemented as an Eclipse plugin. Although it is a rich tool with many features, it can take significant time to set it up properly and is quite feature-heavy. Therefore, we have decided to implement our own lightweight modelling tool as our proof-of-concept. The tool can be accessed from:

@@

The RationalGRL tool is an open-source\footnote{\url{http://www.github.com/marcvanzee/RationalGRL}}, web-based Javascript application, which runs on all modern browsers. The tool is based on our formalization in Section~\ref{sect:formalframework}. Because this formalization is very close to the GRL metamodel, the models can be easily exported to jUCMNav.

When the tool starts, the user is presented with a screen as in Figure~\ref{fig:tool:overview}. This screen shows the palette of elements and links to the left, a canvas on which RationalGRL models can be built, and a pane with details of the currently selected elements on the canvas. Elements and links can be added to the canvas by selecting them on the left and clicking on the canvas, thus capturing the \textsf{INTRO} operation (Section~\ref{sect:formalframework:intro}). This makes it possible to build and modify regular GRL models using the tool.

The details panel can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal element ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). Clicking the ``Answer'' button next to a critical question allows the user to answer the question. As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

Assume that in the example of Figure~\ref{fig:tool:cqdetails} the user selects ``No'' and clicks the ``Answer question'' button. A new argument is then automatically created that attacks the softgoal and the details pane shows the critical question as answered (Figure~\ref{fig:tool:cqeffect}). This is an implementation of a \textsf{DISABLE} algorithm (Section~\ref{sect:formalframework:disable}) similar to Algorithm~\ref{alg:actor-not-relevant}: a new argument is added that attacks the existing argument. It is also possible to attack arguments by adding an ``Argument'' element and an ``Attack'' link manually. Consider, for example, Figure~\ref{fig:tool:argument}. Here, a new argument ``Necessary'', which attacks the previously generated argument based on CQ11, has been added by the user. As the details pane shows, this new argument is not based on a CQ. It is further worth noting that it is possible to provide further a explanation in argument elements, allowing for more fine-grained rationalizations.

@@

The RationalGRL tool computes the acceptability of arguments on the fly. In the example of Figure~\ref{fig:tool:cqeffect}, the original (argument for) softgoal ``Easy to use'' is rejected because it's only attacker ``CQ11 - Redundant'' is accepted. However, if we then attack ``CQ11 - Redundant'' with a new argument ``Necessary'', the original argument for ``Easy to use'' is again accepted because its only attacker is rejected  (Figure~\ref{fig:tool:argument}). Note that when computing the acceptability of arguments, the RationalGRL tool makes the assumption that there are no attack cycles in the model and that hence there is one unique preferred extension (cf. Section~\ref{sect:formalframework:rationalgrl}) -- if the user creates an attack cycle, an error message is shown.

@@

The final operation that has been captured in the RationalGRL tool is \textsf{REPLACE}. For usability purposes, this operation has been implemented differently than in our formal framework (Section~\ref{sect:formalframework:disable}). In the formal framework, a \textsf{REPLACE} introduces a new argument that replaces (and therefore attacks) all previous arguments with the same identifier. Including \textsf{REPLACE} like this in the tool would mean that, when the name of an element is changed, we would need to render all the previous versions of the element on the canvas of the tool. Because this would quickly become very cumbersome, we decided to implement a naming history for each element. For example, in Figure~\ref{fig:tool:overview}, the goal ``Generate cars'' was previously named ``Add cars''. In this way, a history is kept of how an element name was clarified without cluttering the model canvas. Similarly, the decomposition type can be changed in the details pane (Figure~\ref{fig:tool:overview}) without introducing new elements that attack the original ones (cf. Figure~\ref{fig:example-small3}, where the old decomposition type $AND$ of ``Generate traffic'' is shown as an explicit argument).

The tool further implements the RationalGRL to GRL translation (Algorithm~\ref{alg:translation:to-grl}) through an export function. RationalGRL models built in the RationalGRL tool can be exported to the \texttt{.grl} file format, which can be imported in jUCMNav. Two examples of this export are provided in Appendix~\ref{sect:tool-screenshots}. Figure~\ref{fig:tool:figfrompaper} shows the model that was discussed earlier in this paper (Figure~\ref{fig:example-small3}) in the RationalGRL tool. Recall that translating this model to GRL provided us with the model in Figure~\ref{fig:example-small}. This is also what follows from our export: if we export the RationalGRL model and then import the resulting GRL model in jUCMNav, we get the model in Figure~\ref{fig:tool:figfrompaper1}, which is the same as Figure~\ref{fig:example-small}.

@@

>>>>

\section{The RationalGRL Methodology and Tool}
\label{sect:methodology+tool}

In previous sections, we have shown how the RationalGRL framework can capture stakeholder discussions, and how interactions between two types of reasoning, practical reasoning and goal modeling, leads to two interlinked models, RationalGRL and GRL models. The previous section contained a formalization of RationalGRL, which forms the starting point of this section. In this section we clarify how practitioners can actually use the RationalGRL framework by proposing a methodology (\textbf{requirement 4}) and discussing a prototype RationalGRL tool (\textbf{requirement 5}).

@@

We propose the methodology shown in Figure~\ref{fig:rationalgrl-methodology} to develop a RationalGRL model, an earlier version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. %Here we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

@@

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of integrated tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations.

\paragraph{Relation to jUCMNav}

As we discussed in Section~\ref{sect:background:grl}, GRL has a well-documented and well-maintained tool called jUCMNav~\cite{jUCMNav}\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}, which is implemented as an Eclipse plugin. Since this tool is open-source, many researchers have chosen to extend it in various way. However, our tool is not yet integrated into jUCMNav, but consists of a stand-alone web-based version. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is useful by itself as well. The authors of this article have been involved in a large number of empirical studies involving jUCMNav, and our experience is that it can take significant time for participants to setup the Eclipse environment and install the jUCMNav plugin properly. Since our extension does not make use of the many features of jUCMNav, we believe a light-weight web-based version is useful for future empirical research, and it may benefit the community in general. Since RationalGRL is an extension of GRL, our web-based version allows for (limited) GRL modeling. The tool can be accessed from:

@@

\paragraph{Tool overview} The RationalGRL tool is an open-source web-based Javascript application, which runs on all modern browsers. It is based on our formalization in Section~\ref{sect:formalframework} and provides export functionality to jUCMNav, using the translation algorithm (Algorithm~\ref{alg:translation:to-grl}). It contains all of the GRL elements and links, except for beliefs and actors. We have argued in Section~\ref{sect:background:grl} that arguments can be seen as an extension to beliefs, which is the reason why we did not implement them. Actors are also missing, but they will be added in the future.

When the tool starts, the user is presented with a screen as in Figure~\ref{fig:tool:overview}. This screen shows the palette of elements and links (top left pane), a canvas on which RationalGRL models can be built (top right pane), an `Export to GRL' button (bottom left pane), and a details pane of the currently selected elements (bottom right pane). Elements and links can be added to the canvas by selecting them on the left and clicking on the canvas, thus capturing the \textsf{INTRO} operation (Section~\ref{sect:formalframework:intro}). This makes it possible to build and modify GRL models using the tool.

In total there are four different types of details panes, which we now explain in turn.

\paragraph{IE and links details pane} Figure~\ref{fig:tool:overview} shows the details pane for the goal `Generate cars'. The user can change the name of the element by changing the name and clicking `Rename', which will update the naming history on the bottom of the pane. The naming history is (simplified) implemenetation of the \textsf{REPLACE} operation on IEs. If the IE has decomposition links, the user can change the decomposition type as well. Each IE and link has a set of associated critical questions, which can be answered in the details pane by click on the button next to the question.

\paragraph{Critical question details pane} The details panel can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal element ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

Assume that in the example of Figure~\ref{fig:tool:cqdetails} the user selects ``No'' and clicks the ``Answer question'' button. A new argument is then automatically created that attacks the softgoal and the details pane shows the critical question as answered (Figure~\ref{fig:tool:cqeffect}). This is an implementation of a \textsf{DISABLE} algorithm (Section~\ref{sect:formalframework:disable}) similar to Algorithm~\ref{alg:actor-not-relevant}: a new argument is added that attacks the existing argument.

\paragraph{Argument details pane} It is also possible to attack arguments by adding an ``Argument'' element and an ``Attack'' link manually. Consider, for example, Figure~\ref{fig:tool:argument}. Here, a new argument ``Necessary'', which attacks the previously generated argument based on CQ11, has been added by the user. As the details pane shows, this new argument is not based on a CQ. It is further worth noting that it is possible to provide further a explanation in argument elements, allowing for more fine-grained rationalizations.

@@

\paragraph{Argumentation semantics} The RationalGRL tool computes the acceptability of arguments on the fly. In the example of Figure~\ref{fig:tool:cqeffect}, the original (argument for) softgoal ``Easy to use'' is rejected because it's only attacker ``CQ11 - Redundant'' is accepted. However, if we then attack ``CQ11 - Redundant'' with a new argument ``Necessary'', the original argument for ``Easy to use'' is again accepted because its only attacker is rejected  (Figure~\ref{fig:tool:argument}). Note that when computing the acceptability of arguments, the RationalGRL tool makes the assumption that there are no attack cycles in the model and that hence there is one unique preferred extension (cf. Section~\ref{sect:formalframework:rationalgrl}) -- if the user creates an attack cycle, an error message is shown.

@@

\paragraph{Export to jUCMNav} The tool implements the RationalGRL to GRL translation (Algorithm~\ref{alg:translation:to-grl}) through an export function. RationalGRL models built in the RationalGRL tool can be exported to the \texttt{.grl} file format, which can be imported in jUCMNav. Note that in order for the GRL models to be rendered correctly, we recommend using the Graphviz dot\footnote{\url{http://www.graphviz.org/}} as the auto-layout tool (this can be selected under ``Auto layout preferences'' when importing the file in jUCMNav).

Two examples of this export are provided in Appendix~\ref{sect:tool-screenshots}. Figure~\ref{fig:tool:figfrompaper} shows the model that was discussed earlier in this paper (Figure~\ref{fig:example-small3}) in the RationalGRL tool. Recall that translating this model to GRL provided us with the model in Figure~\ref{fig:example-small}. This is also what follows from our export: if we export the RationalGRL model and then import the resulting GRL model in jUCMNav, we get the model in Figure~\ref{fig:tool:figfrompaper1}, which is the same as Figure~\ref{fig:example-small}.

@@

\paragraph{Limitations}
For usability purposes, the \textsf{REPLACE} operation has been implemented differently than in our formal framework (Section~\ref{sect:formalframework:disable}). In the formal framework, a \textsf{REPLACE} introduces a new argument that replaces (and therefore attacks) all previous arguments with the same identifier. Including \textsf{REPLACE} like this in the tool would mean that, when the name of an element is changed, we would need to render all the previous versions of the element on the canvas of the tool. Because this would quickly become very cumbersome, we decided to implement a naming history for each element. For example, in Figure~\ref{fig:tool:overview}, the goal ``Generate cars'' was previously named ``Add cars''. In this way, a history is kept of how an element name was clarified without cluttering the model canvas. Similarly, the decomposition type can be changed in the details pane (Figure~\ref{fig:tool:overview}) without introducing new elements that attack the original ones (cf. Figure~\ref{fig:example-small3}, where the old decomposition type $AND$ of ``Generate traffic'' is shown as an explicit argument).
